Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-17 write-ups/FSE/paper.tex
got the paper to fit in the space for FSE

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16},
based on CPU barrier designs~\cite[ch. 17]{HS08}.
 These barriers employ spinning to ensure threads wait at the barrier until all
threads have arrived, which requires fair
scheduling between workgroups for the barrier to operate correctly.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
%\vspace{-2mm}
\end{figure}

@@

Our new model has no effect on the manner in which traditional GPU
workloads are programmed, such as graphics shaders in OpenGL, Vulkan
and DirectX, and regular data-parallel computational kernels.
%
Furthermore, if our cooperative language extensions are
ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
$\transmit$ annotations are ignored, and $\globalbarrier$ and
$\resizingglobalbarrier$ calls are redirected to an existing global
barrier implementation (so that no resizing occurs in the latter
case)---then a cooperative kernel will behave just like its
non-cooperative counterpart would behave.  A developer
for whom today's occupancy-bound execution model suffices can upgrade
their kernel to a cooperative form and retain backwards-compatibility
with today's GPU platforms by pre-processing away our language
extensions.

@@

Our vision is that support for cooperative kernels will be integrated
in the runtimes of future GPU implementations of OpenCL, with driver
and compiler support for our new primitives.  To experiment with our
ideas on current GPUs, we have developed a prototype that mocks up the
required runtime support via a \emph{megakernel}, and exploits the
occupancy-bound execution model that these GPUs provide to ensure fair
scheduling between workgroups.  We emphasise that an aim of
cooperative kernels is to \emph{avoid} depending on the
occupancy-bound model.  Our prototype exploits this model simply to
allow us to experiment with current GPUs whose proprietary drivers we
cannot change.  We describe the megakernel approach
(\mysec\ref{sec:megakernel}) and detail various aspects of the
scheduler component of our implementation
(\mysec\ref{sec:schedulerimpl}).

@@

\myparagraph{Simplifying assumptions}
%
For ease of implementation, our prototype supports
multitasking a single cooperative kernel with a single non-cooperative
kernel (though the non-cooperative kernel can be invoked many times).
We
assume all variables declared at the root lexical scope of a
cooperative kernel are implicitly $\transmit$-annotated.
%this allows our
%implementation to piggy-back on top of Clang (see below) without
%requiring front-end changes.  
Finally, we require that $\offerkill$,
$\offerfork$ and $\resizingglobalbarrier$ are called from the entry
function of a cooperative kernel.  This allows us to use
$\keyword{goto}$ and $\keyword{return}$ to direct threads into and out
of the kernel.  With these restrictions we can experiment
with interesting irregular algorithms (see
\mysec\ref{sec:experiments}).  A non-mock implementation of
cooperative kernels would not use the megakernel approach, so we do not deem
the engineering effort associated with lifting these restrictions to
be worthwhile.

@@

On receiving a data packet
describing a kernel launch $K$, the GPU scheduler must decide
 how to schedule $K$. Suppose $K$ requests
$N$ workgroups. The scheduler queries the scheduler context.  If
there are at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available workgroups, but a cooperative kernel $K_c$ is
executing. The scheduler can use $K_c$'s channel in the scheduler context to command $K_c$ to provide $N - N_a$ workgroups via $\offerkill$.
Once $N$ workgroups are available,
% the scheduler initialises a
%kernel context $\KC$ for $K$, setting the number of active workgroups
%to $N$ and mapping available workgroups (located in the scheduler
%context) to their new $\KC$ workgroup ids. 
the scheduler then sends $N$
workgroups from the available workgroups to execute kernel $K$
% with
%kernel context $\KC$.
(If the new kernel $K$ is itself a cooperative kernel, the scheduler would be free to provide $K$ with fewer than $N$ active workgroups initially.)

@@

To work around compiler crash and compiler hang bugs in Intel's
current OpenCL drivers, we had to apply some semantics-preserving
changes to some applications.  Driver bugs also meant that we could
not obtain results for some configurations where the non-cooperative
kernel asks for all-but-one workgroups (namely bc, bfs, l-sssp, and
color).  These configurations led to severe failures such as machine
freezes and blue screens.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.\footnote{The right figure has only 3 points per line:
  recall that for bfs there are no runs with ``all-but-one''
  non-cooperative workgroups.}
%
the gather time is higher on the right figure: the rmat22 input graph
is rather wide than deep, so the graph algorithm reaches resizing
barriers less often than for the ecology input of the middle figure
for instance. The scheduler thus has fewer opportunities to collect
workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.

>>>>

The global barrier used in this application is not provided as a GPU
primitive, though previous works have shown that such a global
barrier can be implemented~\cite{XF10,DBLP:conf/oopsla/SorensenDBGR16},
based on CPU barrier designs~\cite[ch. 17]{HS08}.
 These barriers employ spinning to ensure threads wait at the barrier until all
threads have arrived, which requires fair
scheduling between workgroups.% for the barrier to operate correctly.

@@

\begin{lstlisting}
kernel graph_app(global graph *g, 
                 global nodes *n0, global nodes *n1) {
(*@\label{line:cgraph:transmit1}@*)  transmit int level = 0;
(*@\label{line:cgraph:transmit2}@*)  transmit global nodes *in_nodes = n0;
(*@\label{line:cgraph:transmit3}@*)  transmit global nodes *out_nodes = n1;
  while(in_nodes.size > 0) {
(*@\label{line:cgraph:rechunking1}@*)    int tid = get_global_id();
(*@\label{line:cgraph:rechunking2}@*)    int stride = get_global_size();
    for (int i = tid; i < nodes.size; i += stride) {
      process_node(g, in_nodes[i], out_nodes, level);
    }
    swap(&in_nodes, &out_nodes);
(*@\label{line:cgraph:resizing1}@*)    resizing_global_barrier();
    reset(out_nodes);
    level++;
(*@\label{line:cgraph:resizing2}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
%\vspace{-2mm}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\vspace{-4mm}
\end{figure}

@@

Our new model has no effect on the manner in which traditional GPU
workloads (e.g.\ graphics shaders and and regular
data-parallel computational kernels) are programmed.
%
Furthermore, if our cooperative language extensions are
ignored---$\offerkill$ and $\offerfork$ are treated as no-ops,
$\transmit$ annotations are ignored, and $\globalbarrier$ and
$\resizingglobalbarrier$ calls are redirected to an existing global
barrier implementation (so that no resizing occurs in the latter
case)---then a cooperative kernel will behave just like its
non-cooperative counterpart would behave.  
%A developer
%for whom today's occupancy-bound execution model suffices can upgrade
%their kernel to a cooperative form and retain backwards-compatibility
%with today's GPU platforms by pre-processing away our language
%extensions.

@@

Our vision is that support for cooperative kernels will be integrated
in the runtimes of future GPU implementations of OpenCL, with driver
support for our new primitives.  To experiment with our ideas on
current GPUs, we have developed a prototype that mocks up the required
runtime support via a \emph{megakernel}, and exploits the
occupancy-bound execution model that these GPUs provide to ensure fair
scheduling between workgroups.  We emphasise that an aim of
cooperative kernels is to \emph{avoid} depending on the
occupancy-bound model.  Our prototype exploits this model simply to
allow us to experiment with current GPUs whose proprietary drivers we
cannot change.  We describe the megakernel approach
(\mysec\ref{sec:megakernel}) and detail various aspects of the
scheduler component of our implementation
(\mysec\ref{sec:schedulerimpl}).

@@

\myparagraph{Simplifying assumptions}
%
For ease of implementation, our prototype supports multitasking a
single cooperative kernel with a single non-cooperative kernel (though
the non-cooperative kernel can be invoked many times).  
%We assume all
%variables declared at the root lexical scope of a cooperative kernel
%are implicitly $\transmit$-annotated.
%this allows our
%implementation to piggy-back on top of Clang (see below) without
%requiring front-end changes.  
We require that $\offerkill$, $\offerfork$ and
$\resizingglobalbarrier$ are called from the entry function of a
cooperative kernel.  This allows us to use $\keyword{goto}$ and
$\keyword{return}$ to direct threads into and out of the kernel.  With
these restrictions we can experiment with interesting irregular
algorithms (see \mysec\ref{sec:experiments}).  A non-mock
implementation of cooperative kernels would not use the megakernel
approach, so we do not deem the engineering effort associated with
lifting these restrictions to be worthwhile.

@@

On receiving a data packet describing a kernel launch $K$, the GPU
scheduler must decide how to schedule $K$. Suppose $K$ requests $N$
workgroups. The scheduler queries the scheduler context.  If there are
at least $N$ available workgroups, $K$ can be scheduled
immediately. Suppose instead that there are only $N_a < N$ available
workgroups, but a cooperative kernel $K_c$ is executing. The scheduler
can use $K_c$'s channel in the scheduler context to command $K_c$ to
provide $N - N_a$ workgroups via $\offerkill$.  Once $N$ workgroups
are available,
% the scheduler initialises a kernel context $\KC$ for $K$, setting
%the number of active workgroups to $N$ and mapping available
%workgroups (located in the scheduler context) to their new $\KC$
%workgroup ids.
the scheduler then sends $N$ workgroups from the available workgroups
to execute kernel $K$
% with
%kernel context $\KC$.
(If the new kernel $K$ is itself a cooperative kernel, the scheduler
would be free to provide $K$ with fewer than $N$ active workgroups
initially.)

@@

%To work around compiler crash and compiler hang bugs in Intel's
%current OpenCL drivers, we had to apply some semantics-preserving
%changes to some applications.  Driver bugs also meant that we could
%not obtain results for some configurations where the non-cooperative
%kernel asks for all-but-one workgroups (namely bc, bfs, l-sssp, and
%color).  These configurations led to severe failures such as machine
%freezes and blue screens.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.%\footnote{The right figure has only 3 points per line:
  %recall that for bfs there are no runs with ``all-but-one''
  %non-cooperative workgroups.}
%
the gather time is higher on the right figure: the rmat22 input graph
is rather wide than deep, so the graph algorithm reaches resizing
barriers less often than for the ecology input of the middle figure
for instance. The scheduler thus has fewer opportunities to collect
workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.
